---
title: "Cluster Creation Hyrox Dataset - GER"
author: "Jojo Justine"
date: "2025-06-02"
output:
  html_document:
    code_folding: hide
---
```{r}
#install.packages("countrycode")
#install.packages("rsample")
#install.packages("fmsb")
#install.packages("factoextra")
```

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load necessary libraries
library(readr)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(rsample)
library(randomForest)
library(cluster)
library(factoextra)
```
## Germany - Women
## Finding Number of Clusters
**Elbow Method**
```{r}
# Read the CSV file
df <- read.csv("hyrox_results_updated_V4.csv")

# Step 1: Filter Germany female who completed the race
df_filtered <- df %>%
  filter(gender == "female",
         country == "Germany",
         completed == "Yes",
         !is.na(total_time)) %>%
  filter(if_all(starts_with("run_"), ~ .x > 60))  # Exclude unrealistic values

# Step 2: Select run split columns
run_columns <- paste0("run_", 1:8)

# Step 3: Extract run splits and drop NAs
run_data <- df_filtered %>%
  select(all_of(run_columns)) %>%
  drop_na()

# Step 4: Scale data for clustering
run_scaled <- scale(run_data)

# Step 5: Use fviz_nbclust to find optimal number of clusters
fviz_nbclust(run_scaled, kmeans, method = "wss",
             k.max = 10,
             nstart = 10,
             diss = NULL) +
  labs(title = "Elbow Method for Optimal K")

```

**Silhouette Method**
```{r}

fviz_nbclust(run_scaled, kmeans, method = "silhouette",
             k.max = 10, nstart = 10) +
  labs(title = "Silhouette Method for Optimal K",
       x = "Number of Clusters K",
       y = "Average Silhouette Width")
```

Even though Elbow and Silhouette method is suggesting 2 clusters, we are classifying pacing strategies into three performance tiers: top, average, and bottom.

```{r}
# Apply k-means clustering (with k = 3)
set.seed(42)
km <- kmeans(run_scaled, centers = 3, nstart = 25)

# Add cluster labels to the full dataset
df_clustered <- df_filtered %>%
  filter(!is.na(run_1)) %>%
  mutate(Cluster = factor(km$cluster))

# Pivot data for plotting
df_long <- df_clustered %>%
  select(all_of(run_columns), Cluster) %>%
  pivot_longer(-Cluster, names_to = "Run", values_to = "Time")

# Convert Run to numeric index and keep the label
df_long <- df_long %>%
  mutate(Run_num = as.numeric(gsub("run_", "", Run)),
         Run_label = paste0("Run ", Run_num))

ggplot(df_long, aes(x = Run_num, y = Time, color = Cluster, group = Cluster)) +
  stat_summary(fun = mean, geom = "line", size = 1.3) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", size = 1) +
  scale_x_continuous(breaks = 1:8, labels = paste0("Run ", 1:8)) +  # Set correct x-axis
  labs(title = "Pacing Strategy by Cluster (Germany Women - K=3)",
       subtitle = "Based on Run Segment Times",
       y = "Average Run Time (seconds)",
       x = "Run Segment") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11))


```
  
The red group (Cluster 1) is the fastest and most consistent across all run segments, showing low and steady times. The green group (Cluster 2) ran at a moderate pace, with a gradual increase in time. The blue group (Cluster 3) is the slowest overall and showed a steep rise in times early in the race, followed by high and slightly fluctuating times — indicating they struggled more as the race progressed. The regression lines confirm that slower runners tend to slow down more over time.  
  

**Individual Pacing Lines by Cluster**
```{r}
# Assuming df_clustered and run_columns are already defined

# Step 1: Prepare long format for individual trajectories
df_long_individual <- df_clustered %>%
  select(all_of(run_columns), Cluster) %>%
  mutate(AthleteID = row_number()) %>%
  pivot_longer(cols = starts_with("run_"), names_to = "Run", values_to = "Time")

# Step 2: Plot individual lines with transparency, grouped by cluster
ggplot(df_long_individual, aes(x = Run, y = Time, group = AthleteID, color = Cluster)) +
  geom_line(alpha = 0.2) +  # Keep individual pacing lines clearly visible
  stat_summary(aes(group = Cluster), fun = mean, geom = "line", size = 2.5, alpha = 1) +  # Thicker average line
  stat_summary(aes(group = Cluster), fun = mean, geom = "point", size = 3, shape = 21, fill = "white", stroke = 1.2) +  # Highlighted average points
  labs(title = "Individual Pacing Trajectories by Cluster",
       subtitle = "Filtered: Only valid Germany Female finishers | Bold lines = cluster average",
       x = "Run Segment", y = "Run Time (seconds)") +
  scale_y_continuous(limits = c(0, NA), expand = c(0, 0)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11))

```

```{r}
df_clustered %>%
  mutate(Early = rowMeans(select(., run_1:run_4), na.rm = TRUE),
         Late = rowMeans(select(., run_5:run_8), na.rm = TRUE)) %>%
  ggplot(aes(x = Early, y = Late, color = Cluster)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Early vs Late Race Average Pace",
       x = "Average Run Time (Run 1–4)",
       y = "Average Run Time (Run 5–8)") +
  theme_minimal()

```
   
**Clustering Including total_time along with individual run times**
```{r}
# Step 1: Add total_time to run splits
run_columns <- paste0("run_", 1:8)
cluster_data <- df_filtered %>%
  select(all_of(run_columns), total_time) %>%
  drop_na()

# Step 2: Scale all columns (very important!)
cluster_scaled <- scale(cluster_data)

# Step 3: Apply K-means
set.seed(42)
km_with_total <- kmeans(cluster_scaled, centers = 3, nstart = 25)

# Step 4: Assign back to original filtered data
df_clustered <- df_filtered %>%
  filter(!is.na(run_1)) %>%
  mutate(Cluster = factor(km_with_total$cluster))

# Prepare long format for pacing plot
df_long_total <- df_clustered %>%
  select(all_of(run_columns), Cluster) %>%
  pivot_longer(cols = all_of(run_columns), names_to = "Run", values_to = "Time")

df_long_total <- df_long_total %>%
  mutate(Run_num = as.numeric(gsub("run_", "", Run)))

ggplot(df_long_total, aes(x = Run_num, y = Time, color = Cluster, group = Cluster)) +
  stat_summary(fun = mean, geom = "line", size = 1.3) +         # Mean pacing line
  stat_summary(fun = mean, geom = "point", size = 2) +          # Mean points
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", size = 1) +  # Regression line
  scale_x_continuous(breaks = 1:8, labels = paste0("Run ", 1:8)) +
  labs(title = "Pacing Strategy by Cluster (With total_time in Clustering)",
       subtitle = "Average run segment times across clusters",
       y = "Run Time (seconds)", x = "Run Segment") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
df_clustered %>%
  mutate(Early = rowMeans(select(., run_1:run_4), na.rm = TRUE),
         Late = rowMeans(select(., run_5:run_8), na.rm = TRUE)) %>%
  ggplot(aes(x = Early, y = Late, color = Cluster)) +
  geom_point(alpha = 0.6) +
  geom_abline(linetype = "dashed") +
  labs(title = "Early vs Late Run Average by Cluster (total_time included)",
       x = "Average Run 1–4 Time", y = "Average Run 5–8 Time") +
  theme_minimal()

```

```{r}
# STEP 1: Define columns (clustering includes total_time)
event_columns <- c(
  "total_time",
  "run_1", "ski_erg",
  "run_2", "sled_push",
  "run_3", "sled_pull",
  "run_4", "burpee_broad_jumps",
  "run_5", "rowing",
  "run_6", "farmers_carry",
  "run_7", "sandbag_lunges",
  "run_8", "wall_balls"
)

# STEP 2: Prepare data and scale
event_data <- df_filtered %>%
  select(all_of(event_columns)) %>%
  drop_na()

event_scaled <- scale(event_data)

# STEP 3: Perform K-means clustering
set.seed(42)
km_events <- kmeans(event_scaled, centers = 3, nstart = 25)

# STEP 4: Add cluster info
df_clustered_events <- df_filtered %>%
  filter(!is.na(run_1)) %>%
  mutate(Cluster = factor(km_events$cluster))

# STEP 5: Prepare data for plotting (exclude total_time)
plot_columns <- event_columns[event_columns != "total_time"]

df_long_events <- df_clustered_events %>%
  select(all_of(plot_columns), Cluster) %>%
  pivot_longer(cols = all_of(plot_columns), names_to = "Event", values_to = "Time")

# STEP 6: Add event index for regression
df_long_events <- df_long_events %>%
  mutate(Event = factor(Event, levels = plot_columns),
         Event_num = as.numeric(Event))  # For regression

# STEP 7: Plot with regression lines
ggplot(df_long_events, aes(x = Event_num, y = Time, color = Cluster, group = Cluster)) +
  stat_summary(fun = mean, geom = "line", size = 0.8) +  # Cluster average lines
  stat_summary(fun = mean, geom = "point", aes(shape = Cluster), size = 2.5) +  # Cluster average points
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", size = 1) +  # Regression lines per cluster
  scale_x_continuous(breaks = 1:length(plot_columns), labels = plot_columns) +  # Match event labels
  labs(title = "Complete Performance Profile by Cluster",
       subtitle = "With Regression Lines — Runs and Workouts Only",
       y = "Time (seconds)", x = "Event Segment") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11))

```
  
The green group (Cluster 2) is the fastest overall, with the lowest times across most events and a flat regression line, showing consistent performance. The red group (Cluster 1) performed moderately, with times slightly higher than green. The blue group (Cluster 3) was the slowest and most inconsistent, with frequent spikes — especially in event like rburpee broad jumps. Their upward regression line indicates they slowed down more as the race progressed.  


```{r}
# STEP 1: Define full and functional event columns
event_columns <- c(
  "total_time",
  "run_1", "ski_erg",
  "run_2", "sled_push",
  "run_3", "sled_pull",
  "run_4", "burpee_broad_jumps",
  "run_5", "rowing",
  "run_6", "farmers_carry",
  "run_7", "sandbag_lunges",
  "run_8", "wall_balls"
)

functional_events <- c(
  "ski_erg", "sled_push", "sled_pull", "burpee_broad_jumps",
  "rowing", "farmers_carry", "sandbag_lunges", "wall_balls"
)

# STEP 2: Prepare data and scale
event_data <- df_filtered %>%
  select(all_of(event_columns)) %>%
  drop_na()

event_scaled <- scale(event_data)

# STEP 3: Perform K-means clustering
set.seed(42)
km_events <- kmeans(event_scaled, centers = 3, nstart = 25)

# STEP 4: Assign cluster to original dataset
df_clustered_events <- df_filtered %>%
  filter(!is.na(run_1)) %>%
  mutate(Cluster = factor(km_events$cluster))

# STEP 5: Prepare long-format data for functional events only
df_long_functional <- df_clustered_events %>%
  select(all_of(functional_events), Cluster) %>%
  pivot_longer(cols = all_of(functional_events), names_to = "Event", values_to = "Time") %>%
  mutate(Event = factor(Event, levels = functional_events),
         Event_num = as.numeric(Event))  # Needed for regression

# STEP 6: Plot
ggplot(df_long_functional, aes(x = Event_num, y = Time, color = Cluster, group = Cluster)) +
  stat_summary(fun = mean, geom = "line", size = 0.8) +
  stat_summary(fun = mean, geom = "point", aes(shape = Cluster), size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", size = 1) +
  scale_x_continuous(breaks = 1:8, labels = functional_events) +
  labs(title = "Functional Event Performance by Cluster",
       subtitle = "Excludes Running Segments and Total Time",
       y = "Time (seconds)", x = "Functional Workout Event") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11))
```
  
The green group (Cluster 2) was the fastest and most consistent, with low times across all functional workout events and a slightly downward regression line. The red group (Cluster 1) showed moderate performance, with times generally higher than green but lower than blue. The blue group (Cluster 3) was the slowest, especially in events like burpee broad jumps, and sandbag lunges, where their times peaked. Their regression line slopes upward, indicating a gradual decline in performance across functional tasks.