---
title: "Exploratory Data Analysis For Hyrox Dataset"
author: "Jojo Justine"
date: "2025-03-13"
output:
  html_document:
    code_folding: hide
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load necessary libraries
library(tidyverse)
```
## Existing columns

There are total 34 Columns. 
  
  
 event_id, event_name, gender, nationality, age_group, division, total_time, work_time, roxzone_time, 
 run_time, run_1, work_1, roxzone_1, run_2, work_2, roxzone_2, run_3, work_3, roxzone_3, run_4, work_4, 
 roxzone_4, run_5, work_5, roxzone_5, run_6, work_6, roxzone_6, run_7, work_7, roxzone_7, run_8, work_8,
 roxzone_8. 
    
    
 **First few records of the dataset are below.**  
 
```{r}
# Read the CSV file
df <- read.csv("hyrox_results.csv")
#summary(df)
head(df)
```

## Filtering the data
```{r}
countries_to_view <- df %>% count(nationality) %>% filter(n > 500) %>% pull(nationality)
countries_to_view <- countries_to_view[-1]
df_filtered <- df %>% filter(nationality %in% countries_to_view) %>% filter(division == "open")
```

## Adding new columns
Splitting the event_name column into Season, Year, and Event_Venue

```{r}

# Get original column names
original_columns <- names(df)

# Create a copy of 'event_name' to preserve it
df$event_name_original <- df$event_name  

# Split 'event_name' into 'Season', 'Year', and 'Event_Venue'
df <- separate(df, col = event_name, into = c("Season", "Year", "Event_Venue"), sep = " ", extra = "merge", fill = "right")

# Restore 'event_name' and remove temporary column
df$event_name <- df$event_name_original  
df$event_name_original <- NULL  

# Reorder columns: Place 'Season', 'Year', and 'Event_Venue' right after 'event_name'
event_name_index <- which(original_columns == "event_name")  # Find position of event_name
new_column_order <- append(original_columns, c("Season", "Year", "Event_Venue"), after = event_name_index)

# Arrange dataframe in new column order
df <- df[, new_column_order]

# Save the updated dataset
write_excel_csv(df, "hyrox_results_updated.csv")
summary(df)
```
## Null Values

**Finding missing values in various fields.**

```{r}
# Count total number of records
total_records <- nrow(df)
cat("Total number of records:", total_records, "\n")

# Count missing values in the nationality column
missing_nationality <- sum(is.na(df$nationality) | df$nationality == "")
cat("Missing nationality values:", missing_nationality, "\n")


# Count missing values in the age_group column
missing_age <- sum(is.na(df$age_group) | df$age_group == "")
cat("Missing age group values:", missing_age, "\n")

# Count missing values in the division column
missing_division <- sum(is.na(df$division) | df$division == "")
cat("Missing division values:", missing_division, "\n")

# Count Zeroes in total_time column
missing_total_time <- sum(is.na(df$total_time) | df$total_time == "" | df$total_time == "0:00:00")
cat("Zero values in total_time:", missing_total_time, "\n")

```
## Unique Values

**Finding Unique values in age_group field.**

```{r}
# Count occurrences of each age group and convert to data frame
age_group_counts <- as.data.frame(table(df$age_group))

# Print as a table
print(age_group_counts)
```
Below are the assumptions for the age groups.  

| Age Group |Notes                             |
|-----------|----------------------------------|
| MHE       | Men's Health Edition category.   |
| WHE       | Women's Health Edition category. |
| MME       | Men's Masters Edition category.  |
| WWE       | Women's Masters Edition category.|
| X40+      | Mixed 40 and above.              |
  
  
**Finding Unique values in division field.**
```{r}
# Get unique values in the division column
unique_division <- unique(df$division)

# Print unique age groups
print(unique_division)
```


